---
title: "sc_lung_3 lung carcinomas"
output:
  pdf_document: default
  html_document: default
date: "2023-05-23"
---

#BIG NOTE:
#ALL THE GENES ARE IN ENSEMBL NOMENCLATURE, EXCEPT THE MITOCONDRIAL GENES
#THOSE ARE IN THE GENE SYMBOLS NOMENCLATURE

#INSTALLATION:
```{r}
install.packages('MatrixGenerics', version = '3.17')
BiocManager::install("matrixStats")

#installation cibersort:
install.packages("remotes")
remotes::install_github("omnideconv/immunedeconv")
```


#libraries
```{r}
library(Seurat)
library(SeuratData)
library(gridExtra)
library(tidyverse)
library(data.table)
library(SummarizedExperiment)
library(MatrixGenerics)
library(SingleCellExperiment)
library(scds)
library(useful)
library(ggplot2)
suppressMessages(library(immunedeconv))
```

```{r}
#installation of cibersort

#library(devtools)
#install_github("immunogenomics/harmony")
#library("CIBERSORT.R")
#source("CIBERSORT.R")

#intallation of singleR
```

#DETERMINE THE DIRECTORY:
```{r}
knitr::opts_knit$set(root.dir = "C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/")
```


#load the data:
https://www.ebi.ac.uk/gxa/sc/experiments/E-MTAB-6653/downloads 
#1º:Load the cells
```{r}
mtx_cols <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/barcodes.mtx_cols", header = FALSE, sep = "\t")
View(mtx_cols)

barcodes<-mtx_cols$V1
barcodes<-as.data.frame(barcodes,header=FALSE,rownames = FALSE)
file_path_<- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/", "barcodes.tsv")
write.table(barcodes,file_path_,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(barcodes)
```

#2º:Load the genes, convert into tsv and compress into .gz
```{r}
mtx_rows <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/genes.mtx_rows", header = FALSE, sep = "\t")
mtx_rows<-as.data.frame(mtx_rows$V1)
View(mtx_rows)

genes<-mtx_rows$`mtx_rows$V1`
genes<-as.data.frame(genes,header=FALSE,rownames = FALSE)
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/", "genes.tsv")
write.table(genes,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(genes)
```

#3º:TRY TO CONVERT THE ENSEMBL GENES INTO GENE SYMBOL AND DETERMINE THE MITOCONDRIAL GENES:
```{r}
#installations missing:
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("org.Hs.eg.db")

library(Seurat)
library(patchwork)
library(dplyr)
library(biomaRt)
library(org.Hs.eg.db)
library(ggplot2)
library(Matrix)

#input list of the genes in the genes.tsv:
data_genes<-genes$genes

#BiomaRt method:
listEnsembl()
ensembl_<-useEnsembl(biomart = "genes")
datasets<-listDatasets(ensembl_)
ensembl <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
attr<-listAttributes(ensembl)#here we define the reference to which type of conversion type we want to convert the ensembl genes
bm <- getBM(attributes=c("ensembl_gene_id", "external_gene_name"), values=data_genes, mart=ensembl)
#THE TOTAL NUMBER OF GENES: 69299

# Check mitochondrial genes before conversion
print("Mitochondrial genes before conversion:")
print(grep("^MT-", data_genes, value = TRUE))#NO MITOCONDRIAL GENES BEFORE CONVERSION

#DETERMINE THE GENE SYMBOLS:
gene.symbols <- bm$external_gene_name[match(data_genes, bm$ensembl_gene_id)]
features_<-as.matrix(cbind(genes$genes,gene.symbols))
colnames(features_)<-c("ensembl_genes","gene_symbols")

#CREATE A NEW DATAFRAME WITH ENSEMBL GENES AND THE MITOCONDRIAL GENES IN GENE.SYMBOLS:
features<-as.data.frame(features_)
MT_GENES<-print(grep("^MT-",features_, value = TRUE))#13 MITOCONDRIAL GENES AFTER CONVERSION
MT_GENES<-as.list(MT_GENES)
features_d<- features[match(features$gene_symbols, MT_GENES, nomatch = 0) != 0,]

#CREATE THE FINAL FEATURES DATAFRAME:
features_df<-as.data.frame(features_)
#DO ONE BY ONE:
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198695"]<-"MT-ND6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198712"]<-"MT-CO2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198727"]<-"MT-CYB"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198763"]<-"MT-ND2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198786"]<-"MT-ND5"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198804"]<-"MT-CO1"#
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198840"]<-"MT-ND3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198886"]<-"MT-ND4"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198888"]<-"MT-ND1"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198899"]<-"MT-ATP6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198938"]<-"MT-CO3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000212907"]<-"MT-ND4L"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000228253"]<-"MT-ATP8"
View(features_df)
#CREATE THE TSV FILE WITH GENES CALLED FEATURES.TSV
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/", "features.tsv")
write.table(features_df,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
#1823 genes-not converted
#22156 genes from the 23979 were actually converted

######################
#OPTIONAL!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!IF THE WE WANT THE FEATURES TO BE IN GENE #SYMBOLS NOMENCLATURE TO THE ONES THAT CAN DO THAT CONVERTION
###########################
#CREATE A DATAFRAME WITH THE GENE NAMES TO THE ENSEMBL IDS THAT HAD THE CONVERTED RESTPECTIVELY GENE NAMES AND IN CASE THEY DON'T MANTAIN FOR THOSE THE ENSEMBL IDS:
bm$ensembl_gene_id<-ifelse(bm$external_gene_name!="",bm$external_gene_name,bm$ensembl_gene_id )

# View the updated dataframe
print(bm)
features_df<-as.data.frame(bm$ensembl_gene_id)
```


#4º:ALL THE DATASET IN THE ACCORDING FORMATS ACCEPTED BY THE FUNCTION READ10X
```{r}
#gene.column = 1
lung.data <- Read10X(data.dir="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653",gene.column = 1)

lung <- CreateSeuratObject(counts = lung.data, project = "lung", min.cells = 3, min.features = 200)
View(lung)
```


#DETERMINE THE PERCENTAGE OF MITOCONDRIAL GENES ACCORDING TO THE PATTERN:
```{r}
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^mt-") 
#OR
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^MT-") 
VlnPlot(lung,features = c("nFeature_RNA","nCount_RNA","percent.mt"),ncol=3)

```

#OVERVIEW OF THE NFEATURE_RNA AND NCOUNT_RNA AND PERCENT.MT:
```{r}
View(lung@meta.data)
FeatureScatter(lung,feature1 = "nCount_RNA",feature2 = "nFeature_RNA")+geom_smooth(method = "lm")

#COMPARING PLOTS:
plot1 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
plot2
```

#FILTERING:
```{r}
lung<-subset(lung,subset=nFeature_RNA>200 & nFeature_RNA<2500 & percent.mt<5)
#FIND DUBLETS:    [SOME AREAS THAT WERE MORE SEQUENCED, IN DUPLICATES]
lung_sce <- as.SingleCellExperiment(lung)
lung_sce <- cxds_bcds_hybrid(lung_sce, estNdbl=TRUE)#Function to find doublets/multiples in UMI
lung <- AddMetaData(
  object = lung,
  metadata = lung_sce$hybrid_call,
  col.name = "doublet"
)
table(lung_sce$hybrid_call)
lung<- subset(lung, subset = doublet==FALSE)

#FALSE:26215
#TRUE:434 DOUBLETS
```

#3.Normalize data
```{r}
#1º:SEE THE MEANS AND VARIANCE OF THE DATA
means <- sparseMatrixStats::rowMeans2(lung@assays$RNA@counts)
variance <- sparseMatrixStats::rowVars(lung@assays$RNA@counts)
plot(means, variance)
```

#2-A.TRY TO NORMALIZE THE DATA WITH LOG NORMALIZE:
```{r}
lung.norm<- NormalizeData(lung, normalization.method = "LogNormalize")

means <- sparseMatrixStats::rowMeans2(as.matrix(lung.norm@assays$RNA@data))
variance <- sparseMatrixStats::rowVars(lung.norm@assays$RNA@data)
plot(means, variance)
```

#2-B.TRY TO NORMALIZE THE DATA WITH SCTransform:
#NOTE:THE SCTRANSFORM TOOK A LONG TIME AND LEAD THE PC DOWN, SO I OPTED TO GO WITH THE CLASSIC NORMALIZATION:
```{r}
lung_sctranform <- SCTransform(lung,verbose=FALSE)

means_sctranform<- sparseMatrixStats::rowMeans2(lung_sctranform@assays$SCT@data)
variance_sctranform<- sparseMatrixStats::rowVars(lung_sctranform@assays$SCT@data)
plot(means_sctranform, variance_sctranform)

```


#3.SCALING (optional)------------------------------
```{r}
hist(sparseMatrixStats::colSums2(lung.norm@assays$RNA@data), main = "Counts per Cell")
```

#4.IDENTIFY HIGHLY VARIABLE FEATURES (GENES)---------------------------------------
```{r}
nsclc <- FindVariableFeatures(lung.norm)

#WITH LOG NORMALIZE:
to.plot <- data.table(means=rowMeans(nsclc@assays$RNA@data),
                      variance=rowVars(as.matrix(nsclc@assays$RNA@data)),
                      hvg=(rownames(nsclc@assays$RNA@data) %in% VariableFeatures(nsclc)))
####################################
```
#Identify the 10 most highly variable genes:
```{r}
top10<-head(VariableFeatures(nsclc),10)
top100<-head(VariableFeatures(nsclc),100)
print(top10)
print(top100)
```

#Plot variable features with and without labels:
```{r}
plot1<-VariableFeaturePlot(nsclc)
LabelPoints(plot=plot1,points = top10,repel = TRUE)#REPEL=TRUE, SHOWS THET POINTS WELL DISTINGUISH BETWEEN EACH OTHER.
```
##5.SCALING
REMOVE THE DESVIATIONS THAT LEAD TO THE CLUSTER OF THE CELLS, CAUSED BY THE DIFFERENCES ON CELL CYCLE OF THIS CELLS (BIOLOGICAL PROBLEM) OR THE BATCH EFFECT (TECNIQUAL PROBLEM); AND NOT BY THE BIOLOGICAL EFFECT.

```{r}
all.genes<-rownames(nsclc)
nsclc.scaled<-ScaleData(nsclc,features = all.genes)
str(nsclc.scaled)
#notes:
#normalized counts is stored in the data slot
#scaling data stored in scale.data
```
#B.DOWNSTREAM ANALYSIS
#6.Perform Linear dimensionality reduction------------------------------------------------------------------------------
```{r}
#6-A:TO PCA dimensionality reduction
lung<-RunPCA(nsclc.scaled,features = VariableFeatures(object = nsclc.scaled))
print(lung[["pca"]],dims=1:5,nfeatures=10)
DimHeatmap(lung,dims=1,cells=500,balanced = TRUE)
```
```{r}
#6-B:FindNeighbors
#PARAMETERS:
#k.param: defines k for the k-nearest neighbor algorithm (depends- not a right number of k)
#reduction: Reduction to use as input for building the (S)NN
lung <- FindNeighbors(lung, k.param=10, reduction = "pca")
```

```{r}
#6-C:FindClusters
#PARAMETERS:
#RESOLUTION: Value of the resolution parameter, use a value above (below) 1.0 if you want to obtain a larger (smaller) number of communities.
#ex: resolution=0.5- 24 communities; resolution=0.8- 7communities
lung <- FindClusters(lung,resolution = 0.2)
#26 clusters
```

```{r}
#6-D:Plot the clusters:
p <- ggplot(data=as.data.table(lung@meta.data)[,.N,by="seurat_clusters"], mapping = aes(x=seurat_clusters, y=N, fill=seurat_clusters)) +
  geom_bar(stat="identity") +
  labs(x="Cluster Number", y="Number of Cells in that Cluster") +
  theme_classic()
print(p)
```

#UMAP VISUALIZATION:
```{r}
lung <- RunUMAP(lung, dims = 1:10)
#PLOT DIM:
#pt.size: Adjust point size for plotting
#shuffle: Whether to randomly shuffle the order of points. This can be useful for crowded plots if points of interest are being buried. (default is FALSE)
DimPlot(lung, reduction = "umap", pt.size = NULL, shuffle = TRUE) + coord_fixed()
ElbowPlot(lung)
```
#7-CELL ANOTATION-------------------------------------------------------------------------------------------------

#OPTION-A:Method 2-AUTOMATIC ASSIGMENT ANOTATIONS
```{r}
#BiocManager::install("ensembldb")
library(celldex)
ref <- BlueprintEncodeData(ensembl = TRUE)
ref
View(ref)

library(SingleR)
# predict cell types based on the reference
pred <- SingleR(test=as.SingleCellExperiment(lung.norm), ref=ref, labels=ref$label.main)
table(pred$labels)
lung$singleR_annot = pred$labels

#x<-as.data.frame(pred@metadata)

MANUAL_CELL_ANNOTATION<-DimPlot(lung, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
#REFERENCE BASED ASSIGNMENT
REFERENCE_CELL_ANNOTATION<-DimPlot(lung, reduction = "umap", label = TRUE, group.by="singleR_annot",pt.size = 0.5,repel = TRUE) + NoLegend()
MANUAL_CELL_ANNOTATION+REFERENCE_CELL_ANNOTATION
```

#SEE IN WHICH CELLS PLCG1 IS EXPRESSED
```{r}
VlnPlot(lung,features="ENSG00000124181")#expression of PLCG1 between all clusters



EXP_PLCG1<-FeaturePlot(lung, features = "ENSG00000124181")
EXP_EGFR<-FeaturePlot(lung, features = "ENSG00000146648")
#COMPARE THE EXPRESSION OF PLCG1 WITH THE REFERENCE BASED ASSIGNMENT CELL ANNOTATION:
FINAL_PLOT<-REFERENCE_CELL_ANNOTATION+EXP_PLCG1
PLOT_EGFR<-REFERENCE_CELL
pdf(file = "C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/PLCG1_EXP_SC_CELL")
plot(FINAL_PLOT,)
dev.off()

png(filename = "C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/PLCG1_EXP_SC_CELL",) 
plot(FINAL_PLOT)
dev.off()



```
######################
GET THE DATA OF THE COUNTS WITH THE BARCODES ANNOTATED:
```{r}
#1º Retrieve cell type annotations:
cell_annotated <- lung$singleR_annot
PLCG1 <- "ENSG00000124181"

#2º Extract expression values of the gene of interest
expression_values <- FetchData(lung, vars = PLCG1)

#3º Combine cell type annotations and expression values into a table
expression_table <- data.frame(CellType = cell_annotated, Expression = expression_values)

#4º View the expression table
head(expression_table)

#5º Save the file with the expression of PLCG1 across all the cell types annotated automaticall:
save(expression_table,file="expression_table.Rda")

```

######################################################################
#COMPARE THE EXPRESSION BETWEEN THE MAIN CELL TYPES THAT WE SEE THAT PLCG1 IS MORE EXPRESSED-WILCOX-TEST:
######################################################################

```{r}
load("expression_table.Rda")
print(expression_table)
```
```{r}
# 1º: Extract the expression values of the gene for each cell type
expression_endothelial <- expression_table[expression_table$CellType == "Endothelial cells", ]
  
expression_Adipocytes <- expression_table[expression_table$CellType =="Adipocytes", ]
#VS
expression_CD8_T_cells <- expression_table[expression_table$CellType =="CD8+ T-cells", ]

expression_CD4_T_cells <- expression_table[expression_table$CellType =="CD4+ T-cells", ]


# 2º: Perform Wilcoxon test-IT DOESN'T MATTER THE ORDER, BECAUSE THIS TEST DOESN'T TAKE IN ACCOUNT THE DIRECTION OF EXPRESSION
#OPTION_A:expression_endothelial, expression_CD8_T_cells
wilcox_result_A<- wilcox.test(expression_endothelial$ENSG00000124181, expression_CD8_T_cells$ENSG00000124181)
print(wilcox_result_A)

#OPTION_B:expression_endothelial, expression_CD4_T_cells
wilcox_result_B <- wilcox.test(expression_endothelial$ENSG00000124181, expression_CD4_T_cells$ENSG00000124181)
print(wilcox_result_B)

#OPTION_C:expression_Adipocytes, expression_CD8_T_cells
wilcox_result_C<- wilcox.test(expression_Adipocytes$ENSG00000124181, expression_CD8_T_cells$ENSG00000124181)
print(wilcox_result_C)

#OPTION_D:expression_Adipocytes, expression_CD4_T_cells
wilcox_result_D<- wilcox.test(expression_Adipocytes$ENSG00000124181, expression_CD4_T_cells$ENSG00000124181)
print(wilcox_result_D)

```

